---
title: "EnvRtype: Envirotyping for Quantitative Genetics and Plant Breeding"
author: |
  | Germano Costa-Neto^[Maintainer and lead developer, germano.cneto@gmail.com]
  | Fernanda G. Pontes^[Developer, ferpoontes@gmail.com]
date: "Created 12 August 2026 · Revised `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{EnvRtype: Envirotyping for Quantitative Genetics and Plant Breeding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6.5,
  fig.height = 4.2,
  dpi        = 96
)
```

**EnvRtype** is an enviromics toolkit for **genotype-by-environment (GxE)
analysis** and **envirotype-informed genomic prediction**. From geographic
coordinates and date ranges it collects worldwide weather, soil, elevation,
bioclimatic and CMIP6 climate-scenario data, derives agro-meteorological
parameters and a FAO-56 soil water balance, maps daily weather onto crop
developmental stages, and builds environmental covariable matrices, envirotype
typologies, mega-environment clusters and relatedness kernels. It then fits
Bayesian reaction-norm kernel models to predict trained genotypes at untested
locations and planting dates, with coverage diagnostics, spatial mapping and
planting-window optimisation. A simulation engine generates multi-environment
trials and enviromes with known genetic architecture to validate every layer.

This vignette walks through the whole package, section by section, following the
four data layers plus the simulation engine.

```{r setup}
library(EnvRtype)
```

## Why envirotyping

A breeder evaluating 500 hybrids across 30 locations observes that rankings
change from site to site. Classical quantitative genetics treats this
genotype-by-environment interaction as a variance component to be estimated and,
usually, discarded:

$$
y_{ij} = \mu + g_i + e_j + (ge)_{ij} + \varepsilon_{ij}
$$

The term $(ge)_{ij}$ absorbs everything specific to hybrid $i$ in location $j$.
It is estimable only where that combination was actually planted, so it cannot be
transferred to a new site, a new season, or a new planting date. Every untested
combination is a blank.

Envirotyping replaces the *label* $e_j$ with a *description*. Let
$\mathbf{w}_j \in \mathbb{R}^k$ be a vector of $k$ measured environmental
quantities at site $j$ — mean temperature during flowering, accumulated water
deficit during grain fill, radiation at emergence. The interaction then becomes a
function of measurable covariates rather than an unstructured effect:

$$
y_{ij} = \mu + g_i + \mathbf{w}_j^{\top}\boldsymbol{\beta}
       + \sum_{m=1}^{k} g_{im}\, w_{jm} + \varepsilon_{ij}
$$

This is a **reaction norm**: genotype $i$ has its own slope $g_{im}$ against each
environmental gradient $m$. Because $\boldsymbol{\beta}$ and the slopes are
indexed by *covariable*, not by *site*, they carry to any location where
$\mathbf{w}$ can be computed — which, for weather and soil data, is anywhere on
Earth.

That single substitution is what makes prediction into unplanted environments
possible, and it is the organising idea behind every function in this package.

### From covariables to kernels

Fitting $k$ separate slopes per genotype is infeasible when $k$ is in the
hundreds and environments number in the tens. The standard remedy is to work with
**relatedness** rather than with individual coefficients. Define the
environmental kinship

$$
\mathbf{K}_E = \frac{\mathbf{W}\mathbf{W}^{\top}}{\mathrm{tr}(\mathbf{W}\mathbf{W}^{\top})/q}
$$

for $q$ environments, exactly analogous to the genomic relationship matrix of
VanRaden (2008). Two environments are "related" to the degree their covariable
profiles align. The model becomes

$$
\mathbf{y} = \mathbf{1}\mu + \mathbf{Z}_g\mathbf{g} + \mathbf{Z}_e\mathbf{e}
           + \mathbf{g}\!\otimes\!\mathbf{e} + \boldsymbol{\varepsilon},
\qquad
\mathbf{g} \sim N(\mathbf{0}, \mathbf{K}_G\sigma_g^2),
\quad
\mathbf{e} \sim N(\mathbf{0}, \mathbf{K}_E\sigma_e^2)
$$

with the GxE term given a Hadamard product $\mathbf{K}_G \circ \mathbf{K}_E$.
The number of parameters no longer grows with $k$, and prediction at a new site
requires only its row in $\mathbf{K}_E$ — computable from downloaded weather
without a single phenotype.

### Why the growing season must be partitioned

A season-long mean is close to useless. Maize is largely indifferent to heat
during vegetative growth and acutely sensitive during the two weeks around
anthesis; averaging over 120 days dilutes the flowering signal into noise. So
every characterisation function in this package accepts `by.interval = TRUE`
with a `time.window`, producing covariables indexed by *factor × stage* rather
than by factor alone.

The cost is dimensionality: 17 weather factors across 51 three-day intervals
gives 867 indices for 9 environments (Della Coletta et al. 2023). Those indices
are heavily collinear by construction, which is what `env_pca()` exists to
handle.

## How the package is organised

```{r, echo = FALSE, results = "asis"}
cat("
| Layer | Purpose | Key functions |
|---|---|---|
| 1 · Acquire | Pull weather, soil and geodata | `get_weather()`, `get_soil()`, `get_elevation()` |
| 2 · Process | Derive agro-meteorological parameters | `processWTH()`, `water_balance()`, `env_phenology()` |
| 3 · Characterise | Build covariables, typologies, indices | `W_matrix()`, `env_typing()`, `env_pca()` |
| 4 · Model & scan | Kernels, prediction, untested environments | `get_kernel()`, `kernel_model()`, `scan_untested_envs()` |
| Simulate | Ground truth to validate every layer | `sim_met()`, `sim_W()` |
")
```

Layers 1 and 2 require network access, so those chunks are shown but not
evaluated. Everything from layer 3 onward runs on the bundled HELIX maize data
and executes in full.

```{r}
data(maizeWTH)
data(maizeYield)
data(maizeG)

dim(maizeWTH)
dim(maizeYield)
dim(maizeG)
```

`maizeWTH` holds daily weather for five Brazilian sites; `maizeYield` holds grain
yield for 150 hybrids in those five environments; `maizeG` is the genomic
relationship matrix.

---

# 1 · Data acquisition

Weather comes from NASA POWER through **nasapower**; soil from SoilGrids 2.0;
elevation and bioclimate from WorldClim 2; agro-ecological zones from GAEZ v4;
future scenarios from CMIP6.

## Daily weather

```{r, eval = FALSE}
env   <- c("NAIROBI", "NAKURU", "ELDORET")
lat   <- c(-1.28, -0.30, 0.52)
lon   <- c(36.82, 36.08, 35.27)
start <- rep("2020-03-01", 3)
end   <- rep("2020-07-31", 3)

wth <- get_weather(env.id = env, lat = lat, lon = lon,
                   start.day = start, end.day = end)
```

`get_weather()` accepts `parallel = TRUE` with `workers`, chunks long date ranges
via `chunk_size`, and retries failed calls `tries` times with `sleep` seconds
between attempts.

## Hourly weather, day/night partitioned

```{r, eval = FALSE}
wth_h <- get_weather_hourly(env.id = env, lat = lat, lon = lon,
                            start.day = start, end.day = end)
```

This is a separate function rather than an argument on `get_weather()` because
POWER caps hourly requests at 15 parameters, and `T2M_MAX`, `T2M_MIN` and
`FROST_DAYS` are daily aggregates that do not exist hourly. It returns one row
per environment-day with day, night and pre-dawn partitions.

Note the limitation: POWER hourly is MERRA-2 reanalysis, not observation. The
diurnal cycle is modelled.

## Resumable downloads

Large multi-environment downloads fail partway through. The resumable variants
checkpoint to disk:

```{r, eval = FALSE}
wth <- get_weather_resumable(env.id = env, lat = lat, lon = lon,
                             start.day = start, end.day = end,
                             run.id = "kenya2020")

read_progress_log("kenya2020")           # what succeeded, what failed
wth <- restart_from_log("kenya2020")     # resume, retrying failures
```

`get_soil_resumable()` does the same for soil.

## Soil, elevation, bioclimate and zones

```{r, eval = FALSE}
soil <- get_soil(env.id = env, lat = lat, lon = lon)
elev <- get_elevation(env.id = env, lat = lat, lon = lon)
bioc <- get_bioclim(env.id = env, lat = lat, lon = lon, vars = 1:19)
aez  <- get_AEZ(env.id = env, lat = lat, lon = lon)
```

For future projections:

```{r, eval = FALSE}
fut <- get_climate_scenario(env.id = env, lat = lat, lon = lon,
                            scenario = "ssp245")
```

Raster extraction at trial coordinates uses `get_spatial()`.

---

# 2 · Processing

## Agro-meteorological parameters

Raw POWER output gives temperature, radiation and precipitation. What matters
agronomically are derived quantities. `processWTH()` computes all of them:

```{r}
maizeWTH <- processWTH(env.data = maizeWTH, verbose = FALSE)
names(maizeWTH)
```

Three derived columns carry most of the signal:

| Column | Meaning |
|---|---|
| `GDD` | Growing degree days — thermal time |
| `FRUE` | Temperature effect on radiation use efficiency, in `[0, 1]` |
| `PETP` | Precipitation minus potential evapotranspiration; negative = deficit |

```{r}
head(maizeWTH[, c("env", "daysFromStart", "GDD", "FRUE", "PETP")])
```

The three `param_*` functions run individually if you want finer control:

```{r, eval = FALSE}
d <- param_radiation(env.data = maizeWTH, merge = TRUE)
d <- param_atmospheric(env.data = d, Alt = 600)
d <- param_temperature(env.data = d, Tbase1 = 9, Tbase2 = 45,
                       Topt1 = 26, Topt2 = 32)
```

Setting `method = "legacy"` in `processWTH()` reproduces the 2021 behaviour.

## Summarising by environment or stage

```{r}
sm <- summaryWTH(env.data = maizeWTH, env.id = "env",
                 days.id = "daysFromStart",
                 var.id = c("T2M", "PRECTOT", "FRUE"),
                 verbose = FALSE)
head(sm)
```

Split the season into developmental intervals:

```{r}
stages   <- c("VE", "V1_V6", "V6_VT", "VT_R1")
interval <- c(0, 7, 30, 65, 90)

sm_stage <- summaryWTH(env.data = maizeWTH, env.id = "env",
                       days.id = "daysFromStart",
                       var.id = c("FRUE", "PETP"),
                       by.interval = TRUE,
                       time.window = interval,
                       names.window = stages,
                       verbose = FALSE)
head(sm_stage)
```

A season-long mean hides stress concentrated around flowering. Stage resolution
is almost always the more informative view.

## Phenology

`env_phenology()` converts thermal time into developmental stages:

```{r, eval = FALSE}
show_phenology("maize")        # inspect the bundled template
phenology_templates()          # all available crops

ph <- env_phenology(env.data = maizeWTH, env.id = "env",
                    day.id = "daysFromStart",
                    crop = "maize", method = "gdd")
```

Three thermal-time methods are available: `"gdd"` (simple), `"baskerville"`
(sine-wave correction for days crossing the base temperature), and
`"trapezoid"`. Photoperiod sensitivity is enabled with `photoperiod = TRUE`.

## FAO-56 water balance

Combining weather with soil hydraulic properties gives a daily root-zone water
balance:

```{r, eval = FALSE}
wb <- water_balance(env.data = maizeWTH, soil.data = soil,
                    env.id = "env", days.id = "daysFromStart",
                    PREC = "PRECTOT",
                    root.depth = 1000, root.init = 300,
                    dap.root.max = 60)

summary_water_balance(wb, by.interval = TRUE,
                      time.window = interval,
                      names.window = stages)
```

Read the scope limits before drawing agronomic conclusions. This is the FAO-56
**single** coefficient method. It does not separate soil evaporation from
transpiration, does not model runoff — all rainfall is assumed to infiltrate —
and does not model capillary rise or lateral flow. On crusted or steep soils the
no-runoff assumption overestimates stored water and therefore *underestimates*
stress. The default Kc and p values describe a generic 120-day maize crop and are
tabulated starting points, not calibrated values.

---

# 3 · Characterisation

## The environmental covariable matrix

**W** has environments in rows and covariables in columns — the numeric
representation every downstream model needs.

```{r}
W <- W_matrix(env.data = maizeWTH,
              var.id = c("FRUE", "PETP", "T2M_MAX", "T2M_MIN"),
              statistic = "mean",
              verbose = FALSE)
dim(W)
round(W[, 1:4], 2)
```

Stage-resolved, which separates flowering stress from vegetative stress:

```{r}
W_stage <- W_matrix(env.data = maizeWTH,
                    var.id = c("FRUE", "PETP"),
                    by.interval = TRUE,
                    time.window = interval,
                    names.window = stages,
                    verbose = FALSE)
dim(W_stage)
colnames(W_stage)
```

Two covariables across four stages gives eight columns, each with its own
coefficient downstream. `W_matrix()` centres and scales by default; `QC = TRUE`
activates `sd.tol` filtering; `impute` handles missing values.

## Environmental typologies

A covariable says *how much*; an **envirotype** says *what kind*. Where crop
physiology defines meaningful thresholds, discretising at those thresholds is
more informative than a mean, because it preserves the shape of the exposure
distribution rather than collapsing it.

For temperature, the cardinals are $T_b$ (base, below which development stops),
$T_{o1}$ and $T_{o2}$ (the optimum band) and $T_c$ (critical/lethal). Counting
the proportion of days falling in each band gives a frequency profile:

$$
f_{jc} = \frac{1}{n_j}\sum_{d=1}^{n_j}
         \mathbb{1}\{x_{jd} \in \text{class } c\}
$$

Two sites with identical mean temperature can have completely different
profiles — one steady at the optimum, the other oscillating between cold nights
and lethal afternoons. The mean cannot distinguish them; the typology can.

Supply cardinals when the crop physiology is known — maize temperature:
base 0 °C, lower optimum 9 °C, upper optimum 32 °C, lethal 45 °C.
```{r}
card_temp <- c( 9, 32, 45,Inf)

ET <- env_typing(env.data = maizeWTH, var.id = "T2M", env.id = "env",
                 cardinals = card_temp, verbose = FALSE)
head(ET)
```

Without cardinals, classes are cut at observed quantiles:

```{r}
ET_rain <- env_typing(env.data = maizeWTH, var.id = "PRECTOT",
                      env.id = "env", verbose = FALSE)
head(ET_rain)
```

Data-driven envirotype mining is available via `envirotype_mining = TRUE`, which
selects the number of classes by silhouette over `k.range`. Joint typologies
across variables use `joint = TRUE`.

## The envirotype frequency matrix

`T_matrix()` returns environments by envirotype frequency — rows are normalised
compositions:

```{r}
Tm <- T_matrix(env.data = maizeWTH, var.id = "T2M", env.id = "env",
               cardinals = card_temp, verbose = FALSE)
dim(Tm)
```

Because rows are already normalised, pass `is.scaled = TRUE` when feeding a
`T_matrix` to `env_kernel()`.

## Clustering environments into mega-environments

A **mega-environment** is a repeatable group of locations that rank genotypes
similarly. If such groups exist, a breeding programme can select within each
rather than chasing a single genotype that wins everywhere — which usually does
not exist.

`env_cluster()` approximates them by grouping environments with similar
enviromic profiles. The number of clusters is chosen by maximising the average
silhouette width

$$
s(i) = \frac{b(i) - a(i)}{\max\{a(i),\, b(i)\}}
$$

where $a(i)$ is the mean distance from environment $i$ to others in its own
cluster and $b(i)$ the mean distance to the nearest other cluster. Values near 1
mean well-separated groups; values near 0 mean the boundary is arbitrary.

```{r}
cl <- env_cluster(W = W, k.max = 4, seed = 1234, verbose = FALSE)
cl$k
cl$clusters
```

Inspect the silhouette profile before trusting the chosen `k`:

```{r}
cl$silhouette
```

If the best average width is below roughly 0.25, the data do not support discrete
mega-environments and a continuous reaction-norm treatment is more honest than
forcing groups.

The environment kinship implied by the clustering is returned too:

```{r}
round(cl$env.kinship, 2)
```

Forcing a fixed number of groups:

```{r}
cl2 <- env_cluster(W = W, k = 2, seed = 1234, verbose = FALSE)
table(cl2$clusters)
```

Five environments is a small illustration; the method is intended for TPE
surveys with tens to hundreds of sites.

## Which environmental factors drive a grouping

`env_cluster()` is unsupervised — it groups by enviromic profile alone.
`env_target_importance()` is its supervised counterpart: environments are
grouped by **what the breeder cares about** (yield level, heritability,
selection accuracy), and the enviromic data are then asked to explain that
grouping via a Random Forest.

The target `Y` is an environment-level quantity — one value per environment, not
the raw plot data:

```{r}
Y_env <- tapply(maizeYield$value, maizeYield$env, mean)
Y_env <- Y_env[rownames(W)]
round(Y_env, 3)
```

```{r}
if (requireNamespace("randomForest", quietly = TRUE)) {
  imp <- env_target_importance(Y = Y_env, W = W, k = 2, ntree = 200,
                               seed = 1234, verbose = FALSE)
  print(imp$k)
  print(imp$clusters)
  head(imp$importance)
}
```

Variables at the top of `$importance` are candidate drivers of the observed
stratification — the environmental factors worth measuring in future trials, or
worth targeting with stress-tolerance breeding.

A multi-trait target works too. Here environments are grouped by both mean yield
and within-environment variability:

```{r}
if (requireNamespace("randomForest", quietly = TRUE)) {
  Ymat <- cbind(
    mean = Y_env,
    sd   = tapply(maizeYield$value, maizeYield$env, stats::sd)[rownames(W)]
  )
  imp2 <- env_target_importance(Y = Ymat, W = W, k = 2, ntree = 200,
                                seed = 1234, verbose = FALSE)
  head(imp2$importance)
}
```

Use `importance.type = "permutation"` when covariables differ in scale or
cardinality — MeanDecreaseGini is biased towards high-cardinality predictors.

## Multi-year risk profiles and TPE weights

With several years of weather for the same sites, `env_risk_profile()` gives
envirotype frequencies per stage across seasons:

```{r, eval = FALSE}
risk <- env_risk_profile(env.data = long_term_wth,
                         site.id = "env", date.id = "YYYYMMDD",
                         var.id = c("T2M", "PETP"),
                         season.length = 150,
                         crop = "maize", by.stage = TRUE)

tpe_weights(risk, stage = "flowering")
```

Projecting that baseline onto a future scenario:

```{r, eval = FALSE}
project_risk(baseline = risk, scenario = fut)
```

## Copula-based indices

`env_copula()` reparameterises **W** onto copula scales, separating the marginal
behaviour of each covariable from their dependence structure:

```{r}
cop <- env_copula(W = W, index = "pobs")
dim(cop)
```

Options are `"pobs"`, `"joint"`, `"survival"`, `"kendall"` and `"all"`.

## Interval-resolved dissection and PCA

Following Della Coletta et al. (2023), `env_indices()` builds an
environment × (factor × interval) matrix — the paper used 3-day intervals to 151
DAP, giving 51 intervals × 17 factors = 867 indices.

```{r}
idx <- env_indices(env.data = maizeWTH, env.id = "env",
                   day.id = "daysFromStart",
                   var.id = c("FRUE", "PETP", "T2M"),
                   interval = 10L, end.day = 90L,
                   statistic = "mean", verbose = FALSE)
dim(idx)
```

These indices are heavily correlated by construction, which is what PCA is for:

```{r}
pca <- env_pca(W = idx, scale. = TRUE, verbose = FALSE)
head(pca$variance)
```

The result carries `$scores` (environment × PC), `$loadings` (index × PC),
`$variance` (`PC`, `sd`, `prop`, `cum`), the underlying `$prcomp` object,
`$dropped` (constant indices removed before scaling) and `$meta`, which maps
each index back to its factor, interval and day:

```{r}
head(pca$meta)
dim(pca$scores)
dim(pca$loadings)
```

`env_pca()` defaults to `scale. = TRUE` deliberately. The factors sit on wildly
different scales — radiation in MJ/m²/day, precipitation in mm, unitless FRUE in
[0,1] — so an unscaled PCA is dominated by whichever has the largest variance.
Set `scale. = FALSE` for a covariance-matrix PCA. Zero-variance indices are
dropped before scaling rather than silently producing `NaN`.

How many PCs to keep is a judgement, not a fixed number:

```{r, fig.alt = "Scree plot of variance explained by environmental principal components"}
env_pca_scree(pca, verbose = FALSE)
```

```{r, fig.alt = "Environments plotted in the space of the first two principal components"}
env_pca_biplot(pca, pc = c(1, 2), verbose = FALSE)
```

Loadings through the season show *which* factor at *which* growth stage drives a
component:

```{r, fig.alt = "Loess-smoothed PC1 loadings across the growing season by weather factor"}
env_loading_curve(pca, pc = 1, span = 0.4, verbose = FALSE)
```

Correlation structure among indices:

```{r, fig.alt = "Heatmap of correlations among interval-resolved environmental indices"}
env_cor_heatmap(idx, order = "factor", verbose = FALSE)
```

Relating PCs to an external per-environment variable — mean yield, for instance:

```{r}
env_mean <- tapply(maizeYield$value, maizeYield$env, mean)
env_mean <- env_mean[rownames(pca$scores)]

env_pc_associate(pca, y = env_mean, method = "pearson",
                 p.adjust = "BH", verbose = FALSE)
```

A PC that correlates with mean yield identifies an environmental gradient the
crop is actually responding to. Combined with `env_loading_curve()`, which shows
*which factor at which stage* loads on that PC, this is how a vague observation
like "site 3 underperformed" becomes "site 3 accumulated water deficit during
grain fill".

Benjamini–Hochberg adjustment is applied by default because one test is run per
retained PC.

## Soil zoning by Gaussian mixture

```{r, eval = FALSE}
zones <- soil_classification(soil.data = soil)
```

The method follows the standard GMM zoning pipeline with BIC model selection.
Texture fractions are compositional, so a log-ratio transform is applied before
clustering. Read the documentation note: this is not a byte-for-byte
reproduction of Zhang et al. (2025), and SoilGrids carries no heavy metals, so
the cadmium application in that paper cannot be reproduced here.

---

# 4 · Prediction

## Environmental relatedness kernels

Two environments with similar conditions should rank genotypes similarly. A
kernel turns that intuition into a covariance matrix.

**Linear (GB).** The environmental analogue of VanRaden's genomic relationship
matrix:

$$
\mathbf{K}_E = \frac{\mathbf{W}\mathbf{W}^{\top}}
                    {\mathrm{tr}(\mathbf{W}\mathbf{W}^{\top})/q}
$$

The trace normalisation puts the kernel on a scale comparable to $\mathbf{K}_G$,
so variance components from the two sources can be read against each other.

**Gaussian (GK).** Similarity decays nonlinearly with distance:

$$
k(\mathbf{w}_i, \mathbf{w}_j) = \exp\!\left(-h\,
  \frac{\lVert \mathbf{w}_i - \mathbf{w}_j \rVert^2}{q_{50}}\right)
$$

where $q_{50}$ is the median squared Euclidean distance, making the bandwidth
scale-free. Use this when a doubling of stress does more than double the
penalty — common for heat and drought, where damage is threshold-like.

**Arc-cosine deep (DK).** Emulates an infinitely wide neural network with
`deep.layers` hidden layers (Cuevas et al. 2019), recursively:

$$
k^{(l)}(\mathbf{w}_i,\mathbf{w}_j) = \frac{1}{\pi}
  \sqrt{k^{(l-1)}_{ii} k^{(l-1)}_{jj}}
  \left(\sin\theta^{(l-1)} + (\pi - \theta^{(l-1)})\cos\theta^{(l-1)}\right)
$$

No bandwidth to tune, at the cost of interpretability.

```{r}
K_lin <- env_kernel(env.data = W, gaussian = FALSE, verbose = FALSE)
round(K_lin$envCov, 2)
```
```{r}
K_lin <- env_kernel(env.data = W, gaussian = FALSE, verbose = FALSE)
round(K_lin$envCov, 2)
```

The return value has two parts. `envCov` is environment × environment — the one
that goes into the model. `varCov` is covariable × covariable, a diagnostic
showing which covariables move together.

```{r}
K_gau <- env_kernel(env.data = W, gaussian = TRUE, verbose = FALSE)
round(K_gau$envCov, 2)
```

Arc-cosine deep kernels emulate a deep network without a bandwidth parameter:

```{r}
K_deep <- env_kernel(env.data = W, deep.kernel = TRUE, deep.layers = 2,
                     verbose = FALSE)
round(K_deep$envCov, 2)
```

One kernel per developmental stage:

```{r}
K_stage <- env_kernel(env.data = W_stage, stages = stages,
                      gaussian = TRUE, verbose = FALSE)
names(K_stage$envCov)
```

Which kernel performs best is empirical. Compare by cross-validation, not by
inspection.

## Assembling the model

`get_kernel()` combines genomic (`K_G`), environmental (`K_E`) and soil (`K_S`)
kernels into the structure `kernel_model()` expects:

```{r}
K <- get_kernel(K_G = list(G = maizeG),
                K_E = list(W = K_gau$envCov),
                data = maizeYield,
                model = "RNMM",
                env = "env", gid = "gid")
names(K)
```

Four structures are available:

| `model` | Structure | Random terms |
|---|---|---|
| `MM` | Main effects, genomic only | $\mathbf{K}_G$ |
| `MDs` | Main effects plus GxE deviation | $\mathbf{K}_G$, $\mathbf{K}_G \circ \mathbf{Z}_e\mathbf{Z}_e^{\top}$ |
| `RNMM` | Reaction-norm main effects | $\mathbf{K}_G$, $\mathbf{K}_E$ |
| `RNMDs` | Reaction norm plus GxE deviation | $\mathbf{K}_G$, $\mathbf{K}_E$, $\mathbf{K}_G \circ \mathbf{K}_E$ |

The distinction that matters: `MDs` models GxE as an *unstructured* deviation —
estimable only where planted. `RNMM` and `RNMDs` route it through
$\mathbf{K}_E$, which is computable anywhere. Only the latter two can predict
into unplanted environments.

`RNMM` is the usual choice for enviromic prediction — environmental relatedness
enters directly, which is what makes prediction into unplanted environments
possible.

## Fitting

```{r, eval = FALSE}
fit <- kernel_model(y = "value", data = maizeYield, random = K,
                    env = "env", gid = "gid",
                    iterations = 1000, burnin = 200, thining = 10,
                    keep_effects = TRUE)

varcomp_summary(fit)
```

MCMC fitting via **BGGE** is too slow for a vignette build, so these chunks are
not evaluated.

`keep_effects = TRUE` matters more than its name suggests: it retains per-draw
effect chains rather than posterior means. Any downstream scan needs those
chains, or its credible intervals would be fabricated.

For large kernels, eigendecomposition up front speeds up fitting:

```{r, eval = FALSE}
Kd  <- decompose_kernels(K, keep_var = 0.99)
fit <- kernel_model(y = "value", data = maizeYield, random = Kd,
                    env = "env", gid = "gid", keep_effects = TRUE)
```

`truncate_gxe_kernel(random, keep_prop = 0.75)` trims the GxE block;
`undecompose_kernels()` reverses the decomposition.

## Cross-validation

```{r, eval = FALSE}
cv <- kernel_cv(y = "value", data = maizeYield, random = K,
                env = "env", gid = "gid",
                folds = 5, scheme = "cv1", seed = 1234)
```

`cv1` predicts untested genotypes; `cv0` predicts untested environments.

## Multiple chains and clustered models

```{r, eval = FALSE}
fit_mc <- kernel_model_mc(y = "value", data = maizeYield, random = K,
                          env = "env", gid = "gid", n_chains = 3, seed = 1)

cl2 <- cluster_environments(K_gau$envCov, k = 2, method = "hclust")
fit_cl <- kernel_model_clustered(y = "value", data = maizeYield, random = K,
                                 env = "env", gid = "gid",
                                 clusters = cl2, use = "block")
```

## Predicting untested environments

The payoff. Given a fitted model and a covariable vector $\mathbf{w}_*$ for a
site that was never planted, the predicted environmental effect is the
Gaussian-process conditional mean

$$
\hat{e}_* = \mathbf{k}_*^{\top}
            (\mathbf{K}_E + \lambda\mathbf{I})^{-1}\hat{\mathbf{e}}
$$

where $\mathbf{k}_*$ holds the kernel evaluations between the new site and each
training environment. This is kriging in covariable space rather than in
geographic space — two sites 2000 km apart can be neighbours if their climates
match.

**The caveat that matters most.** The formula returns a number for any
$\mathbf{w}_*$, including one far outside anything observed. The arithmetic is
valid; the prediction is not. `scan_untested_envs()` therefore reports where
each new site sits relative to the training envelope, via the Mahalanobis ratio

$$
r_* = \sqrt{(\mathbf{w}_* - \bar{\mathbf{w}})^{\top}
            \mathbf{S}^{-1}(\mathbf{w}_* - \bar{\mathbf{w}})}
      \;\big/\; \max_i r_i
$$

Values above 1 mean the site is further from the training centroid than any
training environment was. Treat those predictions as hypotheses, not forecasts.
```{r, eval = FALSE}
sc <- scan_untested_envs(object = fit,
                         W_train = W, W_new = W_new,
                         K_G = maizeG,
                         level = 0.95,
                         envelope = "flag")

coverage_summary(sc)
```

`envelope` controls extrapolation handling — `"keep"`, `"flag"`, `"mask"` or
`"drop"`. A new environment far outside the training envelope produces a
prediction that is arithmetically valid and scientifically worthless;
`coverage_summary()` tells you which sites those are.

Mapping and tabulating:

```{r, eval = FALSE}
map_scan(sc, coords, type = "coverage")
map_scan(sc, coords, type = "performance", genotypes = "G01")
map_scan(sc, coords, type = "which_won_where")

scan_spatial_table(sc, coords, lon_col = "lon", lat_col = "lat")
```

For large grids, `grid_scan()` chunks the computation to bound memory:

```{r, eval = FALSE}
gs <- grid_scan(fit, W_train = W, W_new = W_grid, K_G = maizeG,
                chunk_size = 500)
```

## Planting dates

The same machinery answers *when* rather than *where*:

```{r, eval = FALSE}
planting_window_table(sc, dates = sowing_dates)
plot_planting_window(sc, dates = sowing_dates, summarise = "genotype")
best_planting_date(sc, dates = sowing_dates, exclude_flagged = TRUE)
```

Combining both — one map panel per candidate date:

```{r, eval = FALSE}
map_scan(sc, coords, date_col = "date", facet_by = "date",
         genotypes = "G01")
```

---

# 5 · Simulation

Every layer above can be validated against known truth. `sim_met()` generates a
multi-environment trial from a specified genomic kinship and environmental
correlation structure.

```{r}
C_env <- sim_met_C(q = 6, min.cor = 0.1, max.cor = 0.8, seed = 42)
round(C_env, 2)
```

```{r}
set.seed(42)
sim <- sim_met(K = maizeG, C = C_env, min.h2 = 0.3, max.h2 = 0.7,
               n_rep = 2, seed = 42)
str(sim$data)
```

The simulator returns `$data` (the phenotypes), `$C_env` (the environmental
correlation used) and `$truth` (the generating parameters).

`sim_W()` then produces a covariable matrix explaining a *known* proportion of
that environmental correlation:

```{r}
W_sim <- sim_W(C_env = sim$C_env, noise = 0.2, n_var = 30,
               collinearity = 0.5, seed = 42, verbose = FALSE)
dim(W_sim)
```

Because the truth is known, the full pipeline can be run on `W_sim` and the
recovered variance components compared against `sim$truth`. That is the
validation loop:

```{r, eval = FALSE}
K_sim <- env_kernel(env.data = W_sim, verbose = FALSE)
Ks    <- get_kernel(K_G = list(G = maizeG),
                    K_E = list(W = K_sim$envCov),
                    data = sim$data, model = "RNMM",
                    env = "env", gid = "gid")
fit_s <- kernel_model(y = "value", data = sim$data, random = Ks,
                      env = "env", gid = "gid")
varcomp_summary(fit_s)      # compare against sim$truth
```

To explore how noise, dimensionality and collinearity affect recovery, sweep a
grid:

```{r, eval = FALSE}
sim_W_grid(C_env = sim$C_env,
           noise = seq(0, 1, 0.25),
           n_var = c(10, 50, 200),
           collinearity = c(0, 0.5, 0.9),
           n_rep = 10, seed = 42)
```

---

# Citation

```{r, eval = FALSE}
citation("EnvRtype")
```

Costa-Neto, G., Galli, G., Carvalho, H. F., Crossa, J., & Fritsche-Neto, R.
(2021). EnvRtype: a software to interplay enviromics and quantitative genomics
in agriculture. *G3: Genes|Genomes|Genetics*, 11(4), jkab040.

# Key references

**Enviromics and prediction** — Costa-Neto et al. (2021, *G3* 11:jkab040);
Costa-Neto, Fritsche-Neto & Crossa (2021, *Heredity* 126:92–106); Costa-Neto et
al. (2023, *G3* 13:jkac313); Cuevas et al. (2019, *G3* 9:2913–2924); Granato et
al. (2018, *G3* 8:3039–3047).

**Characterisation** — Costa-Neto et al. (2023, *Agronomy Journal* 115(5));
Della Coletta et al. (2023, *GENETICS* 224:iyad103).

**Copulas** — Sklar (1959); Genest & Rivest (1993); Salvadori et al. (2007).

**Data sources** — Sparks (2018, *JOSS* 3:1035); Poggio et al. (2021, *SOIL*
7:217–240); Fick & Hijmans (2017, *IJC* 37:4302–4315); O'Neill et al. (2016,
*GMD* 9:3461–3482); FAO & IIASA (2021, GAEZ v4).

**Water balance and phenology** — Allen et al. (1998, FAO-56); Pereira et al.
(2021, *AWM* 243:106466); Doorenbos & Kassam (1979, FAO-33); Zadoks et al.
(1974, *Weed Research* 14:415–421).

**Soil zoning** — Fraley & Raftery (2002, *JASA* 97:611–631); Scrucca et al.
(2016, *R Journal* 8:205–233); Egozcue et al. (2003, *Math. Geol.* 35:279–300).
